### 2018-01-22 paper/ms.tex
changing notation and starting on RNAAS

\documentclass[rnaas]{aastex62}

\pdfoutput=1

\usepackage{lmodern}
\usepackage{microtype}
\usepackage{url}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{natbib}
\usepackage{multirow}
\usepackage{graphicx}
\bibliographystyle{aasjournal}

% ------------------ %
% end of AASTeX mods %
% ------------------ %

% Projects:
\newcommand{\project}[1]{\textsf{#1}}
\newcommand{\kepler}{\project{Kepler}}
\newcommand{\ktwo}{\project{K2}}

% references to text content
\newcommand{\documentname}{\textsl{Note}}
\newcommand{\figureref}[1]{\ref{fig:#1}}
\newcommand{\Figure}[1]{Figure~\figureref{#1}}
\newcommand{\figurelabel}[1]{\label{fig:#1}}
\renewcommand{\eqref}[1]{\ref{eq:#1}}
\newcommand{\Eq}[1]{Equation~(\eqref{#1})}
\newcommand{\eq}[1]{\Eq{#1}}
\newcommand{\eqalt}[1]{Equation~\eqref{#1}}
\newcommand{\eqlabel}[1]{\label{eq:#1}}

% TODOs
\newcommand{\todo}[3]{{\color{#2}\emph{#1}: #3}}
\newcommand{\dfmtodo}[1]{\todo{DFM}{red}{#1}}
\newcommand{\alltodo}[1]{\todo{TEAM}{red}{#1}}
\newcommand{\citeme}{{\color{red}(citation needed)}}

% math
\newcommand{\T}{\ensuremath{\mathrm{T}}}
\newcommand{\dd}{\ensuremath{ \mathrm{d}}}
\newcommand{\unit}[1]{{\ensuremath{ \mathrm{#1}}}}
\newcommand{\bvec}[1]{{\ensuremath{\boldsymbol{#1}}}}
\newcommand{\Gaussian}[3]{\ensuremath{\frac{1}{|2\pi #2|^\frac{1}{2}}
            \exp\left[ -\frac{1}{2}#1^\top #2^{-1} #1 \right]}}

% VECTORS AND MATRICES USED IN THIS PAPER
\newcommand{\Normal}{\ensuremath{\mathcal{N}}}
\newcommand{\mA}{\ensuremath{\bvec{A}}}
\newcommand{\mC}{\ensuremath{\bvec{C}}}
\newcommand{\mS}{\ensuremath{\bvec{\Sigma}}}
\newcommand{\mL}{\ensuremath{\bvec{\Lambda}}}
\newcommand{\vw}{\ensuremath{\bvec{w}}}
\newcommand{\vy}{\ensuremath{\bvec{y}}}
\newcommand{\vt}{\ensuremath{\bvec{\theta}}}
\newcommand{\vm}{\ensuremath{\bvec{\mu}(\bvec{\theta})}}
\newcommand{\vre}{\ensuremath{\bvec{r}}}
\newcommand{\vh}{\ensuremath{\bvec{h}}}
\newcommand{\vk}{\ensuremath{\bvec{k}}}

% typography obsessions
\setlength{\parindent}{3.0ex}

\begin{document}\raggedbottom\sloppy\sloppypar\frenchspacing

\title{%
    Scalable gradients of Gaussian Process models with celerite
}

\author[0000-0002-9328-5652]{Daniel Foreman-Mackey}
\affil{Center for Computational Astrophysics, Flatiron Institute, New York, NY}

\keywords{%
methods: data analysis ---
methods: statistical
}

\section{Abstract}

This research note presents a derivation and implementation of reverse-mode
gradients of the \emph{celerite} algorithm for scalable inference with
Gaussian Process (GP) models.
These gradients can be integrated into existing automatic differentiation
frameworks to provide a scalable and computationally efficient method for
evaluating the gradients of the GP likelihood with respect to the kernel
hyperparameters.
The algorithm derived in this note uses less memory and is more efficient than
versions using automatic differentiation and the computational cost scales
linearly with the number of data points.

\section{Introduction}

Gaussian Processes \citep[GPs][]{Rasmussen:2006} are a class of models used
extensively in the astrophysical literature to model stochastic processes.
The applications are broad-ranging and some example include the time domain
variability of astronomical sources \citep{Brewer:2009, Kelly:2014,
Haywood:2014, Rajpaul:2015, Foreman-Mackey:2017}, data-driven models of light
curves or stellar spectra \citep{Wang:2012, Luger:2016, Czekala:2017}, and the
cosmic microwave background \citep{Bond:1987,Wandelt:2003}.
In all of these applications, the calculation and optimization of the GP
marginalized likelihood function is of central importance and this can often
be of the computational bottleneck.
The details of this model are omitted here (more details can be found in
\citealt{Rasmussen:2006} and \citealt{Foreman-Mackey:2017}), but the key point
is that, for a dataset with $N$ data points, every evaluation of a GP model
requires computation of the inverse\footnote{Or, more specifically, a linear
operator that can left multiply a vector or a matrix by the inverse of this
matrix.} and log-determinant of the $N \times N$ covariance matrix.
The computational cost of these operations scales as $\mathcal{O}(N^3)$ in the
general case, but the \emph{celerite} method was recently introduced in the
astronomical literature to compute the GP likelihood for a class of
one-dimensional models with $\mathcal{O}(N)$ scaling
\citep{Foreman-Mackey:2017}.
The details of these models can be found elsewhere
\citep{Foreman-Mackey:2017}, but the basic idea is that for some covariance
functions, the matrix $K$ can be written as a rank-$J$ semi-separable matrix
\begin{eqnarray}\eqlabel{semi-K}
K &=& \mathrm{diag}(\bvec{a}) + \mathrm{tril}(U\,V^\T) +
    \mathrm{triu}(V\,U^\T)
\end{eqnarray}
where $\bvec{a}$ is a $N$-vector, $U$ and $V$ are $N \times J$ matrices, the
diag function creates a diagonal matrix from a vector, and the tril and triu
extract the strict lower and upper triangular matrices from a dense matrix.
For a matrix of this form, \citet{Foreman-Mackey:2017} derived an algorithm
with $\mathcal{O}(N\,J^2)$ scaling for computing the Cholesky factorization
\begin{eqnarray}
K &=& L\,\mathrm{diag}(\bvec{d})\,L^\T
\end{eqnarray}
where
\begin{eqnarray}\eqlabel{semi-L}
L &=& I + \mathrm{tril}(U\,W^\T)
\end{eqnarray}
for some $N \times J$ matrix $W$.
The algorithm from \citet{Foreman-Mackey:2017} includes an $N-1 \times J$
``preconditioning'' matrix\footnote{This matrix is called $\phi$ by
\citet{Foreman-Mackey:2017}.} that we include with the notation $P$.
The Cholesky factorization algorithm derived by \citet{Foreman-Mackey:2017} is
given is reproduced in the following algorithm:

\medskip
\begin{minipage}{\linewidth}
\textbf{function} \texttt{celerite\_factor}($U$, $P$, $\bvec{d}$, $W$) \\
\hspace*{2em}\textsf{\# If at input $\bvec{d} = \bvec{a}$, $W = V$, and $K$ is
    given by \Eq{semi-K},} \\
\hspace*{2em}\textsf{\# at output $K = L\,L^\T$ where $L$ is given by
    \Eq{semi-L}.} \\
\hspace*{2em}$S \gets$ \texttt{zeros}($J$, $J$) \\
    \hspace*{2em}$\bvec{w}_1 \gets \bvec{w}_1 / d_{1}$ \\
\hspace*{2em}\textbf{for} $n = 2,\ldots,N$:\\
\hspace*{2em}\hspace*{2em}$S \gets \texttt{diag}(\bvec{p}_{n-1})\,[
    S + d_{n-1}\,{\bvec{w}_{n-1}}^\T\,{\bvec{w}_{n-1}}
]\,\texttt{diag}(\bvec{p}_{n-1})$ \\
\hspace*{2em}\hspace*{2em}$d_{n} \gets d_{n} - \bvec{u}_n\,S\,{\bvec{u}_n}^\T$\\
\hspace*{2em}\hspace*{2em}$\bvec{w}_n \gets \left[\bvec{w}_n -
    \bvec{u}_n\,S \right] / d_{n}$\\
    \hspace*{2em}\textbf{return} $\bvec{d}$, $W$, $S$
\end{minipage}
\medskip

\noindent In this algorithm, the \texttt{zeros}$(J,\,K)$ function creates a $J
\times K$ matrix of zeros, the \texttt{diag} function creates a diagonal
matrix from a vector, and $\bvec{x}_n$ indicates a row vector made from the
$n$-th row of the matrix $X$.

Similarly, \citet{Foreman-Mackey:2017} derived a $\mathcal{O}(N\,J)$ algorithm
to apply the inverse of $K$ as shown in the following algorithm:

\medskip
\begin{minipage}{\linewidth}
\textbf{function} \texttt{celerite\_solve}($U$, $P$, $\bvec{d}$, $W$, $Z$) \\
\hspace*{2em}\textsf{\# If at input $Z = Y$, at output $Z = K^{-1}\,Y$.} \\
\hspace*{2em}$F \gets$ \texttt{zeros}($J$, $N_\mathrm{rhs}$) \\
\hspace*{2em}$G \gets$ \texttt{zeros}($J$, $N_\mathrm{rhs}$) \\
\hspace*{2em}\textbf{for} $n = 2,\ldots,N$:\\
\hspace*{2em}\hspace*{2em}$F \gets \texttt{diag}(\bvec{p}_{n-1})\,[F +
    {\bvec{w}_{n-1}}^\T\,\bvec{z}_{n-1}]$\\
\hspace*{2em}\hspace*{2em}$\bvec{z}_n \gets \bvec{z}_n - \bvec{u}_n\,F$\\
\hspace*{2em}\textbf{for} $n = 1,\ldots,N$:\\
\hspace*{2em}\hspace*{2em}$\bvec{z}_n \gets \bvec{z}_n / d_{n}$\\
\hspace*{2em}\textbf{for} $n = N-1,\ldots,1$:\\
\hspace*{2em}\hspace*{2em}$G \gets \texttt{diag}(\bvec{p}_{n})\,[G +
    {\bvec{u}_{n+1}}^\T\,\bvec{z}_{n+1}]$\\
    \hspace*{2em}\hspace*{2em}$\bvec{z}_n \gets \bvec{z}_n - \bvec{w}_n\,G$\\
\hspace*{2em}\textbf{return} $Z$, $F$, $G$
\end{minipage}
\medskip

\citep{Murray:2016} and \citep{Giles:2008}.

\begin{figure}[h!]
\begin{center}
\includegraphics[width=0.9\textwidth]{figure.pdf}
\caption{%
    Scaling.
\figurelabel{figure}}
\end{center}
\end{figure}

%\acknowledgements
%It is a pleasure to thank
%  Patrick Cooper,
%  Boris Leistedt,
%  Bernhard Sch\"olkopf, and
%  Dun Wang
%for helping us understand all of this.

\bibliography{celerite}

\end{document}

### 2018-01-22 src/benchmark.cc

@@ -22,42 +22,51 @@ void run_benchmark (int J, int N) {
   srand(1234);
   vector A(N), Y = vector::Random(N), Z, bA(N), bD(N), bY(N), bZ(N);
   matrix U = matrix::Random(N, J), bU(N, J),
-         V = matrix::Random(N, J), bV(N, J), bW(N, J),
+         V0 = matrix::Random(N, J), V(N, J), bV(N, J), bW(N, J),
          P = matrix::Random(N-1, J), bP(N-1, J);
   Matrix<T, J_comp, J_comp, Options> S(J, J), bS(J, J);
   Matrix<T, J_comp, 1> F(J), G(J), bF(J), bG(J);
   S.setZero();
-  A.setConstant(10*J);
 
   // Likelihood time
-  double strt = get_timestamp();
+  double strt, end, count = 0.0;
 
-  int flag = celerite::factor(U, P, A, V, S);
-  T ll = log(A.array()).sum();
+  strt = get_timestamp();
+  do {
+    V << V0;
+    A.setConstant(10*J);
+    int flag = celerite::factor(U, P, A, V, S);
+    T ll = log(A.array()).sum();
 
-  Z = Y;
-  celerite::solve(U, P, A, V, Z, F, G);
-  ll += Y.transpose() * Z;
+    Z = Y;
+    celerite::solve(U, P, A, V, Z, F, G);
+    ll += Y.transpose() * Z;
 
-  double end = get_timestamp();
+    end = get_timestamp();
+    count += 1.0;
+  } while ((end - strt < 0.7) && (count < 3.0));
   std::cout << sizeof(T) << "," << J_comp << "," << J << "," << N << ",";
-  std::cout << (end - strt) << ",";
+  std::cout << ((end - strt) / count) << ",";
 
   // Grad time
   strt = get_timestamp();
-  bZ = Y;
-  bF.setZero();
-  bG.setZero();
-  celerite::solve_grad(U, P, A, V, Z, F, G, bZ, bF, bG, bU, bP, bD, bW, bY);
-  bY.array() += Z.array();
+  count = 0.0;
+  do {
+    bZ = Y;
+    bF.setZero();
+    bG.setZero();
+    celerite::solve_grad(U, P, A, V, Z, F, G, bZ, bF, bG, bU, bP, bD, bW, bY);
+    bY.array() += Z.array();
 
[…]

### 2018-01-22 src/celerite.h

@@ -6,7 +6,7 @@ template <typename T1, typename T2, typename T3, typename T4, typename T5>
 int factor (
   const Eigen::MatrixBase<T1>& U,  // (N, J)
   const Eigen::MatrixBase<T2>& P,  // (N-1, J)
-  Eigen::MatrixBase<T3>& D,        // (N);    initially set to A
+  Eigen::MatrixBase<T3>& d,        // (N);    initially set to A
   Eigen::MatrixBase<T4>& W,        // (N, J); initially set to V
   Eigen::MatrixBase<T5>& S         // (J, J)
 ) {
@@ -16,22 +16,22 @@ int factor (
 
   // First row
   S.setZero();
-  W.row(0) /= D(0);
+  W.row(0) /= d(0);
 
   // The rest of the rows
   for (int n = 1; n < N; ++n) {
-    // Update S = diag(P) * (S + D*W*W.T) * diag(P)
-    S.noalias() += D(n-1) * W.row(n-1).transpose() * W.row(n-1);
+    // Update S = diag(P) * (S + d*W*W.T) * diag(P)
+    S.noalias() += d(n-1) * W.row(n-1).transpose() * W.row(n-1);
     S.array() *= (P.row(n-1).transpose() * P.row(n-1)).array();
 
-    // Update D = A - U * S * U.T
+    // Update d = a - U * S * U.T
     tmp = U.row(n) * S;
-    D(n) -= tmp * U.row(n).transpose();
-    if (D(n) <= 0.0) return n;
+    d(n) -= tmp * U.row(n).transpose();
+    if (d(n) <= 0.0) return n;
 
-    // Update W = (V - U * S) / D
+    // Update W = (V - U * S) / d
     W.row(n).noalias() -= tmp;
-    W.row(n) /= D(n);
+    W.row(n) /= d(n);
   }
 
   return 0;
@@ -41,15 +41,15 @@ template <typename T1, typename T2, typename T3, typename T4, typename T5, typen
 void factor_grad (
   const Eigen::MatrixBase<T1>& U,   // (N, J)
   const Eigen::MatrixBase<T2>& P,   // (N-1, J)
-  const Eigen::MatrixBase<T3>& D,   // (N)
+  const Eigen::MatrixBase<T3>& d,   // (N)
[…]

### 2018-01-23 paper/ms.tex
adding test of the gradients

@@
 In all of these applications, the calculation and optimization of the GP
-marginalized likelihood function is of central importance and this can often
-be of the computational bottleneck.
+marginalized likelihood function (here we follow the notation of
+\citealt{Foreman-Mackey:2017})
+\begin{eqnarray}\eqlabel{loglike}
+\log \mathcal{L}(\bvec{\theta},\,\bvec{\alpha}) &=&
+    -\frac{1}{2}\,\left[\bvec{y} - \bvec{\mu}_\bvec{\theta}\right]^\T\,
+        {K_\bvec{\alpha}}^{-1}\,\left[\bvec{y}-\bvec{\mu}_\bvec{\theta}\right]
+    -\frac{1}{2}\,\log\det K_\bvec{\alpha} + \mathrm{constant}
+\end{eqnarray}
+is of central importance and this can often be the computational bottleneck.
 The details of this model are omitted here (more details can be found in
@@
 $n$-th row of the matrix $X$.
+Using this factorization, the log-determinant of $K$ is
+\begin{eqnarray}
+\log \det K &=& \sum_{n=1}^N \log d_{n} \quad.
+\end{eqnarray}
 
@@
 
-\citep{Murray:2016} and \citep{Giles:2008}.
+\section{Gradients of GP models using celerite}
+
+The standard method of computing gradients of \Eq{loglike} with respect to the
+parameters $\bvec{\theta}$ and $\bvec{\alpha}$ is the equation
+\citep{Rasmussen:2006}
+\begin{eqnarray}\eqlabel{naive-grad}
+\frac{\dd \log \mathcal{L}}{\dd \alpha_k} &=&
+    \frac{1}{2}\,\mathrm{Tr}\left[
+        \left[
+        \bvec{\tilde{r}}\,\bvec{\tilde{r}}^\T - {K_\bvec{\alpha}^{-1}}
+        \right]
+        \,\frac{\dd K}{\dd \alpha_k}
+    \right]
+\end{eqnarray}
+where
+\begin{eqnarray}
+    \bvec{\tilde{r}} &=&
+        {K_\bvec{\alpha}}^{-1}\,\left[\bvec{y}-\bvec{\mu}_\bvec{\theta}\right]
+    \quad.
+\end{eqnarray}
+Similar equations exist for the parameters $\bvec{\theta}$.
+Even with a scalable method of applying ${K_\bvec{\alpha}}^{-1}$, the
+computational cost of \Eq{naive-grad} scales as $\mathcal{O}(N^2)$.
+This scaling is prohibitive when applying the \emph{celerite} method to large
+datasets and I have not found a simple analytic method of improving this
+scaling for semi-separable matrices.
+However, it was recently demonstrated that substantial computational gains can
+be made by directly differentiating Cholesky factorization algorithms even in
+the general case \citep{Murray:2016}.
+
+Following this reasoning and using the notation from an excellent review of
+matrix gradients \citep{Giles:2008}, I present the reverse-mode gradients of
+the \emph{celerite} method.
+While not yet popular within astrophysics, ``reverse-mode accumulation'' of
+gradients (also known as ``backpropagation'') has revolutionalized the field
+of machine learning over the past few decades (TODO ADD CITATION).
+The review \citep{Giles:2008} provides a thorough overview of these methods
+and the interested reader is directed to that discussion for details and for
+an explanation of the notation.
+
+\medskip
+\begin{minipage}{\linewidth}
+\textbf{function} \texttt{celerite\_factor\_grad}($U$, $P$, $\bvec{d}$, $W$,
+    $S$, $\bar{\bvec{d}}$, $\bar{W}$, $\bar{S}$) \\
+\hspace*{2em}\textbf{for} $n = N,\ldots,2$:\\
+
+\hspace*{2em}$S \gets$ \texttt{zeros}($J$, $J$) \\
+    \hspace*{2em}$\bvec{w}_1 \gets \bvec{w}_1 / d_{1}$ \\
+\hspace*{2em}\textbf{for} $n = 2,\ldots,N$:\\
+\hspace*{2em}\hspace*{2em}$S \gets \texttt{diag}(\bvec{p}_{n-1})\,[
+    S + d_{n-1}\,{\bvec{w}_{n-1}}^\T\,{\bvec{w}_{n-1}}
+]\,\texttt{diag}(\bvec{p}_{n-1})$ \\
+\hspace*{2em}\hspace*{2em}$d_{n} \gets d_{n} - \bvec{u}_n\,S\,{\bvec{u}_n}^\T$\\
+\hspace*{2em}\hspace*{2em}$\bvec{w}_n \gets \left[\bvec{w}_n -
+    \bvec{u}_n\,S \right] / d_{n}$\\
+    \hspace*{2em}\textbf{return} $\bvec{d}$, $W$, $S$
+\end{minipage}
+\medskip
 

### 2018-01-23 src/celerite.h

@@ -65,18 +65,18 @@ void factor_grad (
   for (int n = N-1; n > 0; --n) {
     // Step 6
     bd_ -= W.row(n) * bW_.transpose();
-    bV.row(n).noalias() = bW_;
-    bU.row(n).noalias() = -bW_ * S_;
+    bV.row(n).noalias() += bW_;
+    bU.row(n).noalias() += -bW_ * S_;
     bS_.noalias() -= U.row(n).transpose() * bW_;
 
     // Step 5
-    ba(n) = bd_;
+    ba(n) += bd_;
     bU.row(n).noalias() -= 2.0 * bd_ * U.row(n) * S_;
     bS_.noalias() -= bd_ * U.row(n).transpose() * U.row(n);
 
     // Step 4
     S_ *= P.row(n-1).asDiagonal().inverse();
-    bP.row(n-1).noalias() = (bS_ * S_ + S_.transpose() * bS_).diagonal();
+    bP.row(n-1).noalias() += (bS_ * S_ + S_.transpose() * bS_).diagonal();
 
     // Step 3
     bS_ = P.row(n-1).asDiagonal() * bS_ * P.row(n-1).asDiagonal();
@@ -91,10 +91,10 @@ void factor_grad (
   }
 
   // Finally update the first row.
-  bU.row(0).setZero();
-  bV.row(0).noalias() = bW_;
+  //bU.row(0).setZero();
+  bV.row(0).noalias() += bW_;
   bd_ -= bW_ * W.row(0).transpose();
-  ba(0) = bd_;
+  ba(0) += bd_;
 }
 
 template <typename T1, typename T2, typename T3, typename T4, typename T5>
@@ -152,12 +152,12 @@ void solve_grad (
   Eigen::Matrix<typename T4::Scalar, T4::RowsAtCompileTime, T4::ColsAtCompileTime, T4::IsRowMajor>
     Z_ = Z;
 
-  bY = bZ;
+  bY.array() = bZ.array();
   bU.row(0).setZero();
 
   for (int n = 0; n <= N-2; ++n) {
     // Grad of: Z.row(n).noalias() -= W.row(n) * G;
-    bW.row(n).noalias() = -bY.row(n) * G_.transpose();
+    bW.row(n).noalias() -= bY.row(n) * G_.transpose();
     bG_.noalias() -= W.row(n).transpose() * bY.row(n);
 
     // Inverse of: Z.row(n).noalias() -= W.row(n) * G;
[…]

### 2018-01-23 src/test.cc

@@ -1,0 +1,150 @@
+#include <iostream>
+#include <Eigen/Core>
+#include <sys/time.h>
+
+#include "celerite.h"
+
+using namespace Eigen;
+
+double get_timestamp () {
+  struct timeval now;
+  gettimeofday (&now, NULL);
+  return double(now.tv_usec) * 1.0e-6 + double(now.tv_sec);
+}
+
+template <typename matrix, typename vector>
+typename matrix::Scalar compute_likelihood (
+  const vector& a, const matrix& U, const matrix& V, const matrix& P, const vector& y
+) {
+  typedef typename matrix::Scalar T;
+  const int J = matrix::ColsAtCompileTime;
+  int N = a.rows();
+
+  vector d, z;
+  matrix W;
+  Matrix<T, J, J> S(J, J), bS(J, J);
+  Matrix<T, J, 1> F(J), G(J), bF(J), bG(J);
+
+  S.setZero();
+  d = a;
+  W = V;
+  int flag = celerite::factor(U, P, d, W, S);
+  T ll = log(d.array()).sum();
+
+  z = y;
+  celerite::solve(U, P, d, W, z, F, G);
+  ll += y.transpose() * z;
+
+  return ll;
+}
+
+template <typename matrix, typename vector>
+typename matrix::Scalar compute_grad_likelihood (
+  const vector& a, const matrix& U, const matrix& V, const matrix& P, const vector& y,
+  vector& ba, matrix& bU, matrix& bV, matrix& bP, vector& by
+) {
+  typedef typename matrix::Scalar T;
+  const int J = matrix::ColsAtCompileTime;
+  int N = a.rows();
+
+  vector d, z, bd(N), bz(N);
+  matrix W, bW(N, J);
+  Matrix<T, J, J> S(J, J), bS(J, J);
+  Matrix<T, J, 1> F(J), G(J), bF(J), bG(J);
+
+  S.setZero();
+  d = a;
+  W = V;
+  int flag = celerite::factor(U, P, d, W, S);
+  if (flag) std::cerr << flag << std::endl;
+  T ll = log(d.array()).sum();
+
+  z = y;
+  celerite::solve(U, P, d, W, z, F, G);
+  ll += y.transpose() * z;
+
+  // Seed gradients.
+  bz = y;
+  //by = z;
+  by.setZero();
+  bd.array() = 1.0 / d.array();
+
+  bF.setZero();
+  bG.setZero();
+
+  bU.setZero();
+  bP.setZero();
+  bW.setZero();
+
[…]

### 2018-01-23 src/celerite.h
updating bY instead of overwriting

@@ -150,15 +150,15 @@ void solve_grad (
   Eigen::Matrix<typename T5::Scalar, T5::RowsAtCompileTime, T5::ColsAtCompileTime, T5::IsRowMajor>
     bF_ = bF, F_ = F, bG_ = bG, G_ = G;
   Eigen::Matrix<typename T4::Scalar, T4::RowsAtCompileTime, T4::ColsAtCompileTime, T4::IsRowMajor>
-    Z_ = Z;
+    Z_ = Z, bZ_ = bZ;
 
-  bY.array() = bZ.array();
-  bU.row(0).setZero();
+  // TODO: can replace bZ_ by bY if bY initially assumed to be zero.
+  //bU.row(0).setZero();
 
   for (int n = 0; n <= N-2; ++n) {
     // Grad of: Z.row(n).noalias() -= W.row(n) * G;
-    bW.row(n).noalias() -= bY.row(n) * G_.transpose();
-    bG_.noalias() -= W.row(n).transpose() * bY.row(n);
+    bW.row(n).noalias() -= bZ_.row(n) * G_.transpose();
+    bG_.noalias() -= W.row(n).transpose() * bZ_.row(n);
 
     // Inverse of: Z.row(n).noalias() -= W.row(n) * G;
     Z_.row(n).noalias() += W.row(n) * G_;
@@ -175,21 +175,21 @@ void solve_grad (
 
     // Grad of: g.noalias() += U.row(n+1).transpose() * Z.row(n+1);
     bU.row(n+1).noalias() += Z_.row(n+1) * bG_.transpose();
-    bY.row(n+1).noalias() += U.row(n+1) * bG_;
+    bZ_.row(n+1).noalias() += U.row(n+1) * bG_;
   }
 
-  bW.row(N-1).setZero();
+  //bW.row(N-1).setZero();
 
-  bY.array().colwise() /= d.array();
-  bd.array() -= (Z_.array() * bY.array()).rowwise().sum();
+  bZ_.array().colwise() /= d.array();
+  bd.array() -= (Z_.array() * bZ_.array()).rowwise().sum();
 
   // Inverse of: Z.array().colwise() /= d.array();
   Z_.array().colwise() *= d.array();
 
   for (int n = N-1; n >= 1; --n) {
     // Grad of: Z.row(n).noalias() -= U.row(n) * f;
-    bU.row(n).noalias() -= bY.row(n) * F_.transpose();
[…]

### 2018-01-23 src/test.cc

@@ -65,8 +65,7 @@ typename matrix::Scalar compute_grad_likelihood (
 
   // Seed gradients.
   bz = y;
-  //by = z;
-  by.setZero();
+  by = z;
   bd.array() = 1.0 / d.array();
 
   bF.setZero();
@@ -77,7 +76,6 @@ typename matrix::Scalar compute_grad_likelihood (
   bW.setZero();
 
   celerite::solve_grad(U, P, d, W, z, F, G, bz, bF, bG, bU, bP, bd, bW, by);
-  by.array() += z.array();
 
   bS.setZero();
   ba.setZero();

### 2018-01-23 paper/ms.tex
a few words

@@
 \begin{eqnarray}\eqlabel{semi-K}
-K &=& \mathrm{diag}(\bvec{a}) + \mathrm{tril}(U\,V^\T) +
-    \mathrm{triu}(V\,U^\T)
+K &=& \mathrm{diag}(\bvec{a}) + \mathrm{tril}(\tilde{U}\,\tilde{V}^\T) +
+    \mathrm{triu}(\tilde{V}\,\tilde{U}^\T)
 \end{eqnarray}
-where $\bvec{a}$ is a $N$-vector, $U$ and $V$ are $N \times J$ matrices, the
-diag function creates a diagonal matrix from a vector, and the tril and triu
-extract the strict lower and upper triangular matrices from a dense matrix.
+where $\bvec{a}$ is a $N$-vector, $\tilde{U}$ and $\tilde{V}$ are $N \times J$
+matrices, the diag function creates a diagonal matrix from a vector, and the
+tril and triu extract the strict lower and upper triangular matrices from a
+dense matrix.
 For a matrix of this form, \citet{Foreman-Mackey:2017} derived an algorithm
@@
 \citet{Foreman-Mackey:2017}.} that we include with the notation $P$.
+In this notation,
+\begin{eqnarray}
+\bvec{u}_n &\gets&
+\left[\prod_{k=1}^{n-1}\mathrm{diag}(\bvec{p}_k)^{-1}\right]\,\bvec{u}_n\\
+\bvec{v}_n &\gets&
+\left[\prod_{k=1}^{n-1}\mathrm{diag}(\bvec{p}_k)\right]\,\bvec{v}_n\\
+\bvec{w}_n &\gets&
+\left[\prod_{k=1}^{n-1}\mathrm{diag}(\bvec{p}_k)\right]\,\bvec{w}_n\\
+\end{eqnarray}
+and in all that follows, I will use these transformed
+variables\footnote{\citet{Foreman-Mackey:2017} indicate the transformed
+variables using a tilde, but that notation will be inconsistent later.}.
+
 The Cholesky factorization algorithm derived by \citet{Foreman-Mackey:2017} is
@@
 \hspace*{2em}\textbf{for} $n = N,\ldots,2$:\\
+\hspace*{2em}\hspace*{2em}$\bar{d}$
 

### 2018-01-24 paper/ms.tex
words and inplace grad algorithm

--- 2018-01-23 paper/ms.tex
@@
 \begin{eqnarray}\eqlabel{semi-K}
-K &=& \mathrm{diag}(\bvec{a}) + \mathrm{tril}(U\,V^\T) +
-    \mathrm{triu}(V\,U^\T)
+K &=& \mathrm{diag}(\bvec{a}) + \mathrm{tril}(U\,{V}^\T) +
+    \mathrm{triu}({V}\,{U}^\T)
 \end{eqnarray}
-where $\bvec{a}$ is a $N$-vector, $U$ and $V$ are $N \times J$ matrices, the
-diag function creates a diagonal matrix from a vector, and the tril and triu
-extract the strict lower and upper triangular matrices from a dense matrix.
+where $\bvec{a}$ is a $N$-vector, ${U}$ and ${V}$ are $N \times J$
+matrices, the diag function creates a diagonal matrix from a vector, and the
+tril and triu extract the strict lower and upper triangular matrices from a
+dense matrix.
 For a matrix of this form, \citet{Foreman-Mackey:2017} derived an algorithm
@@
 \citet{Foreman-Mackey:2017}.} that we include with the notation $P$.
+The preconditioned variables were indicated with a tilde by
+\citet{Foreman-Mackey:2017}, but to simplify the notation, we will only refer
+to the pre-conditioned variables in all that follows.
+
 The Cholesky factorization algorithm derived by \citet{Foreman-Mackey:2017} is
@@
 \textbf{function} \texttt{celerite\_factor\_grad}($U$, $P$, $\bvec{d}$, $W$,
-    $S$, $\bar{\bvec{d}}$, $\bar{W}$, $\bar{S}$) \\
+    $S$, $\bar{S}$, $\bar{\bvec{a}}$, $\bar{V}$) \\
+\hspace*{2em}\textsf{\# At input $\bar{\bvec{a}} = \bar{\bvec{d}}$ and
+    $\bar{V} = \bar{W}$}\\
+    \hspace*{2em}$\bar{\bvec{v}}_N \gets \bar{\bvec{v}}_N / d_N$\\
 \hspace*{2em}\textbf{for} $n = N,\ldots,2$:\\
+\hspace*{2em}\hspace*{2em}$\bar{a}_n \gets \bar{a}_n -
+    \bvec{w}_n\,{\bar{\bvec{v}}_n}^\T$\\
+\hspace*{2em}\hspace*{2em}$\bar{\bvec{u}}_n \gets - [\bar{\bvec{v}}_n +
+    2\,\bar{a}_n\,\bvec{u}_n]\,S$\\
+\hspace*{2em}\hspace*{2em}$\bar{S} \gets \bar{S} -
+    {\bvec{u}_n}^\T\,[\bar{\bvec{v}}_n + \bar{a}_n\,\bvec{u}_n]$\\
 
-\hspace*{2em}$S \gets$ \texttt{zeros}($J$, $J$) \\
-    \hspace*{2em}$\bvec{w}_1 \gets \bvec{w}_1 / d_{1}$ \\
-\hspace*{2em}\textbf{for} $n = 2,\ldots,N$:\\
-\hspace*{2em}\hspace*{2em}$S \gets \texttt{diag}(\bvec{p}_{n-1})\,[
-    S + d_{n-1}\,{\bvec{w}_{n-1}}^\T\,{\bvec{w}_{n-1}}
-]\,\texttt{diag}(\bvec{p}_{n-1})$ \\
-\hspace*{2em}\hspace*{2em}$d_{n} \gets d_{n} - \bvec{u}_n\,S\,{\bvec{u}_n}^\T$\\
-\hspace*{2em}\hspace*{2em}$\bvec{w}_n \gets \left[\bvec{w}_n -
-    \bvec{u}_n\,S \right] / d_{n}$\\
-    \hspace*{2em}\textbf{return} $\bvec{d}$, $W$, $S$
+\hspace*{2em}\hspace*{2em}$\bar{\bvec{p}}_{n-1} \gets \mathrm{diag}(\bar{S}\,S
+    \,\mathrm{diag}(\bvec{p}_{n-1})^{-1} + \mathrm{diag}(\bvec{p}_{n-1})^{-1}\,
+    S\,\bar{S})$\\
+\hspace*{2em}\hspace*{2em}$\bar{S} \gets \mathrm{diag}(\bvec{p}_{n-1})\,
+    \bar{S}\,\mathrm{diag}(\bvec{p}_{n-1})$\\
+\hspace*{2em}\hspace*{2em}$\bar{d}_{n-1} \gets \bar{d}_{n-1} +
+    \bvec{w}_{n-1}\,\bar{S}\,{\bvec{w}_{n-1}}^\T$\\
+\hspace*{2em}\hspace*{2em}$\bar{\bvec{v}}_{n-1} \gets \bar{\bvec{v}}_{n-1}
+    / d_{n-1} + \bvec{w}_{n-1}\,[\bar{S} + \bar{S}^\T]$\\
+
+\hspace*{2em}\hspace*{2em}$S \gets \mathrm{diag}(\bvec{p}_{n-1})^{-1}\,S\,
+    \mathrm{diag}(\bvec{p}_{n-1})^{-1}
+    - d_{n-1}\,{\bvec{w}_{n-1}}^\T\,\bvec{w}_{n-1}$\\
+
+\hspace*{2em}$\bar{a}_1 \gets \bar{a}_1 -
+    \bar{\bvec{v}}_1\,{\bvec{w}_1}^\T$\\
+
+\hspace*{2em}\textbf{return} $\bar{U}$, $\bar{P}$, $\bar{\bvec{a}}$, $\bar{V}$
 \end{minipage}

### 2018-01-24 src/benchmark.cc

@@ -53,15 +53,20 @@ void run_benchmark (int J, int N) {
   count = 0.0;
   do {
     bZ = Y;
+    bY = Z;
+    bD.array() = 1.0 / A.array();
+
     bF.setZero();
     bG.setZero();
-    celerite::solve_grad(U, P, A, V, Z, F, G, bZ, bF, bG, bU, bP, bD, bW, bY);
-    bY.array() += Z.array();
 
-    bD.array() = 1.0 / A.array();
-    bW.setZero();
+    bU.setZero();
+    bP.setZero();
+
+    celerite::solve_grad(U, P, A, V, Z, F, G, bZ, bF, bG, bU, bP, bD, bW, bY);
+
     bS.setZero();
-    celerite::factor_grad(U, P, A, V, S, bD, bW, bS, bA, bU, bV, bP);
+    celerite::factor_grad(U, P, A, V, S, bS, bU, bP, bD, bW);
+
     end = get_timestamp();
     count += 1.0;
   } while ((end - strt < 0.7) && (count < 3.0));

### 2018-01-24 src/celerite.h

@@ -45,34 +45,26 @@ void factor_grad (
   const Eigen::MatrixBase<T1>& W,   // (N, J)
   const Eigen::MatrixBase<T4>& S,   // (J, J)
 
-  const Eigen::MatrixBase<T3>& bd,  // (N)
-  const Eigen::MatrixBase<T1>& bW,  // (N, J)
   const Eigen::MatrixBase<T4>& bS,  // (J, J)
 
-  Eigen::MatrixBase<T5>& ba,        // (N)
-  Eigen::MatrixBase<T6>& bU,        // (N, J)
-  Eigen::MatrixBase<T6>& bV,        // (N, J)
-  Eigen::MatrixBase<T7>& bP         // (N-1, J)
+  Eigen::MatrixBase<T5>& bU,        // (N, J)
+  Eigen::MatrixBase<T6>& bP,        // (N-1, J)
+  Eigen::MatrixBase<T7>& ba,        // (N)
+  Eigen::MatrixBase<T5>& bV         // (N, J)
 ) {
   int N = U.rows();
 
   // Make local copies of the gradients that we need.
-  typename T3::Scalar bd_ = bd(N-1);
   Eigen::Matrix<typename T4::Scalar, T4::RowsAtCompileTime, T4::ColsAtCompileTime, T4::IsRowMajor> bS_ = bS, S_ = S;
-  Eigen::Matrix<typename T1::Scalar, 1, T1::ColsAtCompileTime> bW_ = bW.row(N-1) / d(N-1);
   Eigen::Matrix<typename T1::Scalar, T1::ColsAtCompileTime, 1> bSWT;
+
+  bV.array().colwise() /= d.array();
 
   for (int n = N-1; n > 0; --n) {
     // Step 6
-    bd_ -= W.row(n) * bW_.transpose();
-    bV.row(n).noalias() += bW_;
-    bU.row(n).noalias() += -bW_ * S_;
-    bS_.noalias() -= U.row(n).transpose() * bW_;
-
-    // Step 5
-    ba(n) += bd_;
-    bU.row(n).noalias() -= 2.0 * bd_ * U.row(n) * S_;
-    bS_.noalias() -= bd_ * U.row(n).transpose() * U.row(n);
+    ba(n) -= W.row(n) * bV.row(n).transpose();
+    bU.row(n).noalias() -= (bV.row(n) + 2.0 * ba.row(n) * U.row(n)) * S_;
[…]

### 2018-01-24 src/test.cc

@@ -78,9 +78,9 @@ typename matrix::Scalar compute_grad_likelihood (
   celerite::solve_grad(U, P, d, W, z, F, G, bz, bF, bG, bU, bP, bd, bW, by);
 
   bS.setZero();
-  ba.setZero();
-  bV.setZero();
-  celerite::factor_grad(U, P, d, W, S, bd, bW, bS, ba, bU, bV, bP);
+  ba = bd;
+  bV = bW;
+  celerite::factor_grad(U, P, d, W, S, bS, bU, bP, ba, bV);
 
   return ll;
 }
@@ -136,13 +136,11 @@ void run_test (int N) {
 
   std::cout << max_error << std::endl;
   std::cout << max_error_name << std::endl;
-  //std::cout << bU2.col(0).transpose() << std::endl;
-  //std::cout << (bU.col(0) - bU2.col(0)).transpose() << std::endl;
 }
 
 int main ()
 {
-  run_test<double, 2>(10);
+  run_test<double, 5>(10);
 
   return 0;
 }

### 2018-01-25 paper/ms.tex
words

@@
 \citep{Foreman-Mackey:2017}.
-The details of these models can be found elsewhere
-\citep{Foreman-Mackey:2017}, but the basic idea is that for some covariance
-functions, the matrix $K$ can be written as a rank-$J$ semi-separable matrix
-\begin{eqnarray}\eqlabel{semi-K}
-K &=& \mathrm{diag}(\bvec{a}) + \mathrm{tril}(U\,{V}^\T) +
-    \mathrm{triu}({V}\,{U}^\T)
-\end{eqnarray}
-where $\bvec{a}$ is a $N$-vector, ${U}$ and ${V}$ are $N \times J$
-matrices, the diag function creates a diagonal matrix from a vector, and the
-tril and triu extract the strict lower and upper triangular matrices from a
-dense matrix.
-For a matrix of this form, \citet{Foreman-Mackey:2017} derived an algorithm
-with $\mathcal{O}(N\,J^2)$ scaling for computing the Cholesky factorization
-\begin{eqnarray}
-K &=& L\,\mathrm{diag}(\bvec{d})\,L^\T
-\end{eqnarray}
-where
-\begin{eqnarray}\eqlabel{semi-L}
-L &=& I + \mathrm{tril}(U\,W^\T)
-\end{eqnarray}
-for some $N \times J$ matrix $W$.
-The algorithm from \citet{Foreman-Mackey:2017} includes an $N-1 \times J$
-``preconditioning'' matrix\footnote{This matrix is called $\phi$ by
-\citet{Foreman-Mackey:2017}.} that we include with the notation $P$.
-The preconditioned variables were indicated with a tilde by
-\citet{Foreman-Mackey:2017}, but to simplify the notation, we will only refer
-to the pre-conditioned variables in all that follows.
 
-The Cholesky factorization algorithm derived by \citet{Foreman-Mackey:2017} is
-given is reproduced in the following algorithm:
+The details of the \emph{celerite} method can be found in
+\citet{Foreman-Mackey:2017} and I will not repeat them here.
+The only difference in notation here is that all the matrices to which I refer
+are the ``pre-conditioned'' matrices that are indicated with a tilde by
+\citet{Foreman-Mackey:2017}, but the tilde is not included here for
+simplicity.
+I will also use the symbol $P$ to mean the $N-1 \times J$ pre-conditioning
+matrix that was called $\phi$ by \citet{Foreman-Mackey:2017}.
+The Cholesky factorization algorithm derived by \citet[][their
+Equation~46]{Foreman-Mackey:2017} is given is listed below in the following
+algorithm:
 
@@
 \textbf{function} \texttt{celerite\_factor}($U$, $P$, $\bvec{d}$, $W$) \\
-\hspace*{2em}\textsf{\# If at input $\bvec{d} = \bvec{a}$, $W = V$, and $K$ is
-    given by \Eq{semi-K},} \\
-\hspace*{2em}\textsf{\# at output $K = L\,L^\T$ where $L$ is given by
-    \Eq{semi-L}.} \\
+\hspace*{2em}\textsf{\# At input $\bvec{d} = \bvec{a}$ and $W = V$} \\
 \hspace*{2em}$S \gets$ \texttt{zeros}($J$, $J$) \\
@@
 \end{eqnarray}
-
 Similarly, \citet{Foreman-Mackey:2017} derived a $\mathcal{O}(N\,J)$ algorithm
-to apply the inverse of $K$ as shown in the following algorithm:
+to apply the inverse of $K$ as follows (Equations~47 and~48 in
+\citealt{Foreman-Mackey:2017}):
 
@@
 \textbf{function} \texttt{celerite\_solve}($U$, $P$, $\bvec{d}$, $W$, $Z$) \\
-\hspace*{2em}\textsf{\# If at input $Z = Y$, at output $Z = K^{-1}\,Y$.} \\
+\hspace*{2em}\textsf{\# At input $Z = Y$} \\
 \hspace*{2em}$F \gets$ \texttt{zeros}($J$, $N_\mathrm{rhs}$) \\
-\hspace*{2em}$G \gets$ \texttt{zeros}($J$, $N_\mathrm{rhs}$) \\
 \hspace*{2em}\textbf{for} $n = 2,\ldots,N$:\\
@@
 \hspace*{2em}\hspace*{2em}$\bvec{z}_n \gets \bvec{z}_n / d_{n}$\\
+\hspace*{2em}$G \gets$ \texttt{zeros}($J$, $N_\mathrm{rhs}$) \\
 \hspace*{2em}\textbf{for} $n = N-1,\ldots,1$:\\
@@
 
+Using the notation from \citet{Giles:2008} and performing some tedious
+algebra, the reverse-mode implementation of the \texttt{celerite\_factor}
+function is found to be:
+
 \medskip
 \begin{minipage}{\linewidth}
-\textbf{function} \texttt{celerite\_factor\_grad}($U$, $P$, $\bvec{d}$, $W$,
+\textbf{function} \texttt{celerite\_factor\_rev}($U$, $P$, $\bvec{d}$, $W$,
     $S$, $\bar{S}$, $\bar{\bvec{a}}$, $\bar{V}$) \\
@@
     {\bvec{u}_n}^\T\,[\bar{\bvec{v}}_n + \bar{a}_n\,\bvec{u}_n]$\\
-
 \hspace*{2em}\hspace*{2em}$\bar{\bvec{p}}_{n-1} \gets \mathrm{diag}(\bar{S}\,S
@@
     / d_{n-1} + \bvec{w}_{n-1}\,[\bar{S} + \bar{S}^\T]$\\
-
 \hspace*{2em}\hspace*{2em}$S \gets \mathrm{diag}(\bvec{p}_{n-1})^{-1}\,S\,
@@
     - d_{n-1}\,{\bvec{w}_{n-1}}^\T\,\bvec{w}_{n-1}$\\
-
 \hspace*{2em}$\bar{a}_1 \gets \bar{a}_1 -
     \bar{\bvec{v}}_1\,{\bvec{w}_1}^\T$\\
-
 \hspace*{2em}\textbf{return} $\bar{U}$, $\bar{P}$, $\bar{\bvec{a}}$, $\bar{V}$
@@
 \medskip
+
+\noindent Similarly, the reverse-mode function corresponding to
+\texttt{celerite\_solve} is:
+
+\medskip
+\begin{minipage}{\linewidth}
+\textbf{function} \texttt{celerite\_solve\_rev}($U$, $P$, $\bvec{d}$, $W$,
+    $Z$, $F$, $G$, $\bar{F}$, $\bar{G}$, $\bar{Y}$) \\
+\hspace*{2em}\textsf{\# At input $\bar{Y} = \bar{Z}$}\\
+\hspace*{2em}\textbf{for} $n = 1,\ldots,N-1$:\\
+\hspace*{2em}\hspace*{2em}$\bar{\bvec{w}}_n \gets
+    - \bar{\bvec{y}}_n\,G^\T$\\
+\hspace*{2em}\hspace*{2em}$\bar{G} \gets \bar{G} - {\bvec{w}_n}^\T \,
+    \bar{\bvec{y}}_n$\\
+\hspace*{2em}\hspace*{2em}$\bvec{z}_n \gets \bvec{z}_n + \bvec{w}_n\,G$\\
+\hspace*{2em}\hspace*{2em}$G \gets \mathrm{diag}(\bvec{p}_n)^{-1}\,G$\\
+\hspace*{2em}\hspace*{2em}$\bar{\bvec{p}}_n \gets \mathrm{diag}(
+    \bar{G}\,G^\T)$\\
+\hspace*{2em}\hspace*{2em}$\bar{G} \gets
+    \mathrm{diag}(\bvec{p}_n)\,\bar{G}$\\
+\hspace*{2em}\hspace*{2em}$G \gets G - {\bvec{u}_{n+1}}^\T\,\bvec{z}_{n+1}$\\
+\hspace*{2em}\hspace*{2em}$\bar{\bvec{u}}_{n+1} \gets
+    \bvec{z}_{n+1}\,\bar{G}^\T$\\
+\hspace*{2em}\hspace*{2em}$\bar{\bvec{y}}_{n+1} \gets
+    \bvec{u}_{n+1}\,\bar{G}$\\
+\hspace*{2em}\textbf{for} $n = 1,\ldots,N$:\\
+\hspace*{2em}\hspace*{2em}$\bar{\bvec{y}}_n \gets \bar{\bvec{y}}_n/d_n$\\
+\hspace*{2em}\hspace*{2em}$\bar{d}_n \gets -
+    \bvec{z}_n\,{\bar{\bvec{y}}_n}^\T$\\
+\hspace*{2em}\textbf{for} $n = N,\ldots,2$:\\
+\hspace*{2em}\hspace*{2em}$\bar{\bvec{u}}_n \gets \bar{\bvec{u}}_n
+    - \bar{\bvec{y}}_n\,F^\T$\\
+\hspace*{2em}\hspace*{2em}$\bar{F} \gets \bar{F} - {\bvec{u}_n}^\T \,
+    \bar{\bvec{y}}_n$\\
+\hspace*{2em}\hspace*{2em}$F \gets \mathrm{diag}(\bvec{p}_{n-1})^{-1}\,F$\\
+\hspace*{2em}\hspace*{2em}$\bar{\bvec{p}}_{n-1} \gets \bar{\bvec{p}}_{n-1} +
+    \mathrm{diag}(\bar{F}\,F^\T)$\\
+\hspace*{2em}\hspace*{2em}$\bar{F} \gets
+    \mathrm{diag}(\bvec{p}_{n-1})\,\bar{F}$\\
+\hspace*{2em}\hspace*{2em}$F \gets F - {\bvec{w}_{n-1}}^\T\,\bvec{z}_{n-1}$\\
+\hspace*{2em}\hspace*{2em}$\bar{\bvec{w}}_{n-1} \gets \bar{\bvec{w}}_{n-1} +
+    \bvec{z}_{n-1}\,\bar{F}^\T$\\
+\hspace*{2em}\hspace*{2em}$\bar{\bvec{y}}_{n-1} \gets \bar{\bvec{y}}_{n-1} +
+    \bvec{w}_{n-1}\,\bar{F}$\\
+\hspace*{2em}\textbf{return} $\bar{U}$, $\bar{P}$, $\bar{\bvec{d}}$,
+    $\bar{W}$, $\bar{Y}$
+\end{minipage}
+\medskip
+
+\noindent A reference C++ implementation of this algorithm can be found online
+TODO ADD ZENODO LINK and \Figure{figure} shows the performance of this
+implementation.
+
+\section{Discussion}
+
+This research note presents the algorithms needed to efficiently compute
+gradients of GP models applied to large datasets using the \emph{celerite}
+method.
+These developments increase the performance of inference methods based on
+\emph{celerite} and improve the convergence properties of non-linear
+optimization routines.
+Furthermore, the derivation of reverse-mode accumulation algorithms for
+\emph{celerite} allow its integration into popular model building and
+automatic differentiation libraries like Stan \citep{Carpenter:2015},
+TensorFlow \citep{Abadi:2016}, and others.
 

### 2018-01-25 src/celerite.h

@@ -140,7 +140,6 @@ void solve_grad (
     Z_ = Z, bZ_ = bZ;
 
   // TODO: can replace bZ_ by bY if bY initially assumed to be zero.
-  //bU.row(0).setZero();
 
   for (int n = 0; n <= N-2; ++n) {
     // Grad of: Z.row(n).noalias() -= W.row(n) * G;
@@ -164,8 +163,6 @@ void solve_grad (
     bU.row(n+1).noalias() += Z_.row(n+1) * bG_.transpose();
     bZ_.row(n+1).noalias() += U.row(n+1) * bG_;
   }
-
-  //bW.row(N-1).setZero();
 
   bZ_.array().colwise() /= d.array();
   bd.array() -= (Z_.array() * bZ_.array()).rowwise().sum();

### 2018-01-25 src/benchmark.cc
updated benchmark

@@ -44,7 +44,7 @@ void run_benchmark (int J, int N) {
 
     end = get_timestamp();
     count += 1.0;
-  } while ((end - strt < 0.7) && (count < 3.0));
+  } while ((end - strt < 0.7) || (count < 3.0));
   std::cout << sizeof(T) << "," << J_comp << "," << J << "," << N << ",";
   std::cout << ((end - strt) / count) << ",";
 
@@ -69,7 +69,7 @@ void run_benchmark (int J, int N) {
 
     end = get_timestamp();
     count += 1.0;
-  } while ((end - strt < 0.7) && (count < 3.0));
+  } while ((end - strt < 0.7) || (count < 3.0));
 
   std::cout << ((end - strt) / count);
   std::cout << std::endl;

### 2018-01-25 paper/ms.tex
caption and acks

@@
 
-\section{Abstract}
+\section{Introduction}
 
@@
 
-\section{Introduction}
-
-Gaussian Processes \citep[GPs][]{Rasmussen:2006} are a class of models used
+GPs \citep{Rasmussen:2006} are a class of models used
 extensively in the astrophysical literature to model stochastic processes.
@@
 one-dimensional models with $\mathcal{O}(N)$ scaling
-\citep{Foreman-Mackey:2017}.
+\citep{Ambikasaran:2015, Foreman-Mackey:2017}.
 
@@
 gradients (also known as ``backpropagation'') has revolutionalized the field
-of machine learning over the past few decades (TODO ADD CITATION).
+of machine learning over the past few decades \citep[for
+example][]{LeCun:2015}.
 The review \citep{Giles:2008} provides a thorough overview of these methods
@@
 
-\begin{figure}[h!]
+It is a pleasure to thank
+Eric Agol,
+Sivaram Ambikasaran, and
+Victor Minden
+for conversations that inspired this work.
+A reference implementation of these algorithms, benchmarks, and tests can be
+found at \url{https://github.com/dfm/celerite-grad}.
+
+\begin{figure}[htbp]
 \begin{center}
@@
 \caption{%
-    Scaling.
+    The computational scaling of the algorithms prsented in this note.
+    \emph{(top row)}: The computational cost of computing \eq{loglike} as a
+    function of the number of datapoints ($N$; left) and the model complexity
+    ($J$; right).
+    \emph{(bottom row)}: The cost of computing the gradient of \eq{loglike}
+    with respect to the vector $\bvec{a}$ and the matrices $U$, $V$, and $P$.
+    In the left panels, each line corresponds to a different value of $J$ as
+    indicated in the legend (with $J$ increasing from bottom to top).
+    Similarly, in the right panels, the lines correspond to different values
+    of $N$ increasing from bottom to top.
 \figurelabel{figure}}
@@
 
-%\acknowledgements
-%It is a pleasure to thank
-%  Patrick Cooper,
-%  Boris Leistedt,
-%  Bernhard Sch\"olkopf, and
-%  Dun Wang
-%for helping us understand all of this.
-
 \bibliography{celerite}

### 2018-01-28 paper/ms.tex
typos from agol

@@
 
-This research note presents a derivation and implementation of reverse-mode
-gradients of the \emph{celerite} algorithm for scalable inference with
-Gaussian Process (GP) models.
-These gradients can be integrated into existing automatic differentiation
-frameworks to provide a scalable and computationally efficient method for
-evaluating the gradients of the GP likelihood with respect to the kernel
-hyperparameters.
+This research note presents a derivation and implementation of efficient and
+scalable gradient computation using the \emph{celerite} algorithm for Gaussian
+Process (GP) modeling.
+The algorithms are derived in a ``reverse accumulation'' or
+``backpropagation'' framework and they can be easily integrated into existing
+automatic differentiation frameworks to provide a scalable method for
+evaluating the gradients of the GP likelihood with respect to all input
+parameters.
 The algorithm derived in this note uses less memory and is more efficient than
@@
 extensively in the astrophysical literature to model stochastic processes.
-The applications are broad-ranging and some example include the time domain
+The applications are broad-ranging and some examples include the time domain
 variability of astronomical sources \citep{Brewer:2009, Kelly:2014,
@@
 \end{eqnarray}
-is of central importance and this can often be the computational bottleneck.
-The details of this model are omitted here (more details can be found in
-\citealt{Rasmussen:2006} and \citealt{Foreman-Mackey:2017}), but the key point
-is that, for a dataset with $N$ data points, every evaluation of a GP model
-requires computation of the inverse\footnote{Or, more specifically, a linear
-operator that can left multiply a vector or a matrix by the inverse of this
-matrix.} and log-determinant of the $N \times N$ covariance matrix.
+is generally the computational bottleneck.
+The details of these models are omitted here (see \citealt{Rasmussen:2006} and
+\citealt{Foreman-Mackey:2017} for details), but the key point is that, for a
+dataset with $N$ data points, every evaluation of a GP model requires
+computation of the inverse\footnote{Or, more specifically, a linear operator
+that can left multiply a vector or a matrix by the inverse of this matrix.}
+and log-determinant of the $N \times N$ covariance matrix $K_\bvec{\alpha}$.
 The computational cost of these operations scales as $\mathcal{O}(N^3)$ in the
@@
 \citet{Foreman-Mackey:2017} and I will not repeat them here.
-The only difference in notation here is that all the matrices to which I refer
-are the ``pre-conditioned'' matrices that are indicated with a tilde by
-\citet{Foreman-Mackey:2017}, but the tilde is not included here for
-simplicity.
-I will also use the symbol $P$ to mean the $N-1 \times J$ pre-conditioning
-matrix that was called $\phi$ by \citet{Foreman-Mackey:2017}.
+The only difference in notation is that all the matrices in what follows are
+the ``pre-conditioned'' matrices that are indicated with a tilde by
+\citet{Foreman-Mackey:2017}.
+The tilde is not included here for simplicity and to improve the readability
+of the algorithms.
+I will also use the symbol $P$ for the $(N-1) \times J$ pre-conditioning matrix
+that was called $\phi$ by \citet{Foreman-Mackey:2017}.
 The Cholesky factorization algorithm derived by \citet[][their
-Equation~46]{Foreman-Mackey:2017} is given is listed below in the following
-algorithm:
+Equation~46]{Foreman-Mackey:2017} is as follows:
 
@@
 $n$-th row of the matrix $X$.
+The computational cost of this algorithm scales as $\mathcal{O}(N\,J^2)$.
 Using this factorization, the log-determinant of $K$ is
@@
 
+\noindent The empirical scaling of these algorithms is shown in
+\Figure{figure}.
+
 \section{Gradients of GP models using celerite}
 
-The standard method of computing gradients of \Eq{loglike} with respect to the
-parameters $\bvec{\theta}$ and $\bvec{\alpha}$ is the equation
+It is standard practice to make inferences using \Eq{loglike} by optimizing or
+choosing a prior and sampling with respect to $\bvec{\theta}$ and
+$\bvec{\alpha}$.
+Many numerical inference methods (like non-linear optimization or Hamiltonian
+Monte Carlo) can benefit from efficient calculation of the gradient of
+\Eq{loglike} with respect to the parameters.
+The standard method of computing these gradients uses the identity
 \citep{Rasmussen:2006}
@@
 the \emph{celerite} method.
-While not yet popular within astrophysics, ``reverse-mode accumulation'' of
-gradients (also known as ``backpropagation'') has revolutionalized the field
-of machine learning over the past few decades \citep[for
-example][]{LeCun:2015}.
+While not yet popular within astrophysics, ``reverse accumulation'' of
+gradients (also known as ``backpropagation'') has recently revolutionized the
+field of machine learning \citep[see][for example]{LeCun:2015} by enabling the
+non-linear optimization of models with large numbers of parameters.
 The review \citep{Giles:2008} provides a thorough overview of these methods
@@
 
-Using the notation from \citet{Giles:2008} and performing some tedious
-algebra, the reverse-mode implementation of the \texttt{celerite\_factor}
-function is found to be:
+Using the notation from \citet{Giles:2008}, after some tedious
+algebra, the reverse accumulation function corresponding to
+\texttt{celerite\_factor} is found to be:
 
@@
     $\bar{V} = \bar{W}$}\\
+\hspace*{2em}$\bar{U} \gets \mathrm{zeros}(N,\,J)$\\
+\hspace*{2em}$\bar{P} \gets \mathrm{zeros}(N-1,\,J)$\\
 \hspace*{2em}$\bar{\bvec{v}}_N \gets \bar{\bvec{v}}_N / d_N$\\
@@
 
-\noindent Similarly, the reverse-mode function corresponding to
+\noindent Similarly, the reverse accumulation function for to
 \texttt{celerite\_solve} is:
@@
 \hspace*{2em}\textsf{\# At input $\bar{Y} = \bar{Z}$}\\
+\hspace*{2em}$\bar{U} \gets \mathrm{zeros}(N,\,J)$\\
+\hspace*{2em}$\bar{P} \gets \mathrm{zeros}(N-1,\,J)$\\
+\hspace*{2em}$\bar{\bvec{d}} \gets \mathrm{zeros}(N)$\\
+\hspace*{2em}$\bar{W} \gets \mathrm{zeros}(N,\,J)$\\
 \hspace*{2em}\textbf{for} $n = 1,\ldots,N-1$:\\
@@
 \noindent A reference C++ implementation of this algorithm can be found online
-TODO ADD ZENODO LINK and \Figure{figure} shows the performance of this
+\citep{Foreman-Mackey:2018} and \Figure{figure} shows the performance of this
 implementation.
@@
 optimization routines.
-Furthermore, the derivation of reverse-mode accumulation algorithms for
+Furthermore, the derivation of reverse accumulation algorithms for
 \emph{celerite} allow its integration into popular model building and
@@
 
+\acknowledgments
 It is a pleasure to thank
@@
 A reference implementation of these algorithms, benchmarks, and tests can be
-found at \url{https://github.com/dfm/celerite-grad}.
+found at \url{https://github.com/dfm/celerite-grad} and
+\citet{Foreman-Mackey:2018}.
 
@@
 \caption{%
-    The computational scaling of the algorithms prsented in this note.
-    \emph{(top row)}: The computational cost of computing \eq{loglike} as a
-    function of the number of datapoints ($N$; left) and the model complexity
+    The empirical computational scaling of the algorithms presented in this
+    note.
+    \emph{(top row)}: The cost of computing \eq{loglike} as a
+    function of the number of data points ($N$; left) and the model complexity
     ($J$; right).
@@
     of $N$ increasing from bottom to top.
+    In both cases, the theoretical scaling is $\mathcal{O}(N,J^2)$.
 \figurelabel{figure}}

### 2018-01-28 paper/ms.tex
angus citation and typo

@@
 This research note presents a derivation and implementation of efficient and
-scalable gradient computation using the \emph{celerite} algorithm for Gaussian
+scalable gradient computations using the \emph{celerite} algorithm for Gaussian
 Process (GP) modeling.
@@
 variability of astronomical sources \citep{Brewer:2009, Kelly:2014,
-Haywood:2014, Rajpaul:2015, Foreman-Mackey:2017}, data-driven models of light
-curves or stellar spectra \citep{Wang:2012, Luger:2016, Czekala:2017}, and the
-cosmic microwave background \citep{Bond:1987,Wandelt:2003}.
+Haywood:2014, Rajpaul:2015, Foreman-Mackey:2017, Angus:2018}, data-driven
+models of light curves or stellar spectra \citep{Wang:2012, Luger:2016,
+Czekala:2017}, and the cosmic microwave background
+\citep{Bond:1987,Wandelt:2003}.
 In all of these applications, the calculation and optimization of the GP

### 2018-01-30 paper/ms.tex
title
submitted version

@@
 \title{%
-    Scalable gradients of Gaussian Process models with celerite
+    Scalable backpropagation for Gaussian Processes using celerite
 }
@@
 dataset with $N$ data points, every evaluation of a GP model requires
-computation of the inverse\footnote{Or, more specifically, a linear operator
-that can left multiply a vector or a matrix by the inverse of this matrix.}
-and log-determinant of the $N \times N$ covariance matrix $K_\bvec{\alpha}$.
+computing the log-determinant and multiplying a vector by the inverse of
+%\footnote{Or, more specifically, a linear operator
+%that can left multiply a vector or a matrix by the inverse of this matrix.}
+the $N \times N$ covariance matrix $K_\bvec{\alpha}$.
 The computational cost of these operations scales as $\mathcal{O}(N^3)$ in the
@@
 Similarly, \citet{Foreman-Mackey:2017} derived a $\mathcal{O}(N\,J)$ algorithm
-to apply the inverse of $K$ as follows (Equations~47 and~48 in
-\citealt{Foreman-Mackey:2017}):
+to apply the inverse of $K$ (i.e.\ compute $Z = K^{-1}\,Y$) as follows
+(Equations~47 and~48 in \citealt{Foreman-Mackey:2017}):
 
@@
 
+\newpage
 \section{Discussion}
@@
 \begin{center}
-\includegraphics[width=0.9\textwidth]{figure.pdf}
+\includegraphics[width=0.8\textwidth]{figure.pdf}
 \caption{%
